---
title: "chess_data"
author: "Gabriel Tarriba"
date: "5/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(reticulate)
```

## Analysis of chess results

In this document I analyse a dataset with the results of over 20,000 chess games played on [lichess](https://lichess.org/), one of the most popular platforms to play chess online. The dataset is on [Kaggle](https://www.kaggle.com/datasnaek/chess) and was uploaded by user [Mitchell J](https://www.kaggle.com/datasnaek). 

I will perform Exploratory Data Analysis as well as some simple Regression Analysis/Machine Learning to find out:

1. What is in the data 
2. How large white's advantage is
3. How rating difference relates to game length (in number of turns)
3. How the rating difference between the two players predicts whether the game will have a decisive result (i.e. not end in a draw)
5. What predicts wether a game will end by checkmate, resignation or time out

```{r cars}
# Load the dataset and look at its structure
df <- read.csv("games.csv")
str(df)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
